My wonderful paper

### 2024-09-12 _manuscript/_tex/index.tex
the toy example + some notesg

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

@@

{[}background{]}

@@

\begin{itemize}
\item
  describe/ define what analysis plan is, expectation (outcome and plan)
\item
  why it is useful to be explicit about expectation?

  \begin{itemize}
  \tightlist
  \item
    expectations can be used to formulate unit tests, which helps to
    divide the ``result universe''
  \end{itemize}
\end{itemize}

\section{Examples}\label{sec-examples}

@@

(these two paragraphs may go to a new section \emph{discussion}) A
result universe constructed in this example can be presented to analysts
to answer the what-if questions raised in the data analysis. What if the
sample size is increased? What if temperature is added to the model?
What would the results expect to be changed when the correlation
structure is different? For analysts, expectations can be used as unit
tests to divide the result universe, which allows them to understand the
impact of each factor on the outcome expectation.

The result universe also provides a holistic view of how the results
obtained by the analysts are situated in all possible results. This can
be seen as a direct towards trustworthy data analysis, where the
audience of the analysis to exercise their own cognitive model
\citep{grolemund_cognitive_2014} to evaluate the results reported.

>>>>

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

@@

{[}background - In a data analysis workflow{]}

For expert analysts, the expectation of the analysis process is often
implicit. They may have a mental model of the analysis process and the
expected results. However, for non-expert analysts, the expectation of
the analysis process may not be clear. They may not know what to expect
from the analysis process, or what to look for when an analysis process
fails the expectation. An analysis plan can help to make the expectation
of the analysis process explicit. It can help to guide the analysis
process, to communicate and compare the analysis process to others, and
to evaluate the analysis process.

@@

\begin{itemize}
\tightlist
\item
  describe/ define what analysis plan is, expectation (outcome and plan)

  \begin{itemize}
  \tightlist
  \item
    expectations can be used to formulate unit tests, they can be used
    to guide to diagnose in the analysis process: data issue,
    unreasonable expectation -- expectation is ``wrong''
  \item
    some expectations can be about data: e.g.~check temporal gaps in
    time series data. Some others are statistical checks: sample size,
    outliers, model specification, or checks based on domain knowledge:
    include temperature in the mortality \textasciitilde{} PM10 model.
  \end{itemize}
\end{itemize}

Q: Whether we should formulate these concept with math notation?? A:
only if it helps

An analysis plan is a set of analysis steps plus their expected
outcomes. An analysis plan consists of two parts: outcome expectation
and plan expectation. Outcome expectation is the expected result of the
analysis, for example, a significant p-value or a coefficient estimate
within a certain range. Plan expectation is the expected results from
the intermediate steps of the analysis, for example, the expected
distribution of the data, or the expected correlation structure between
variables.

While there is typically one major result expectation in an analysis,
multiple plan expectations can be formulated as unit tests to guide the
analysis process. These plan expectations serve as road marks to test,
at various stages of the analysis, whether the results are within
expectation. When the outcome is out of expectation, these plan
expectations can serve as unit tests to diagnose the analysis process.

\begin{itemize}
\tightlist
\item
  why it is useful to be explicit about expectation?

  \begin{itemize}
  \tightlist
  \item
    with simulation, the unit tests can be used to construct the
    ``result universe''
  \item
    With formulated expectation, one can generate multiple version of
    the dataset for teaching statistics
  \end{itemize}
\end{itemize}

When the expectation is formulated, we can also construct values fulfill
or not fulfill the expectation. With simulation, this allows to
construct the set of all possible results that can be obtained from the
analysis process, which we called the result universe. With such a
result universe, we can see how the results obtained by the analysts are
situated in all possible results. When accompanied by the result
universe, it is clearer how different factors in the analysis affects
the main results. This can be seen as a direct towards trustworthy data
analysis, where the audience of the analysis to exercise their own
cognitive model \citep{grolemund_cognitive_2014} to evaluate the results
reported.

Let's think about a 5-day step count. You make a resolution to walk on
average 5000 steps a day (your expectation) and using an app to record
your step count. After 5 days, the app tells you've walked on average
8000 steps.

It is easy to come up with reasons why an 8000 average step is resulted
based on common sense:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  you may run a 10k on day 1, resulting a high step count on the day
  (outlier on the right).
\item
  you left your phone at home on day 3, resulting a zero or minimal step
  count on the day (outlier on the left).
\item
  you may realise the step count may increase since you were in a hiking
  trip in the last five days (average shift).
\end{enumerate}

Based on these reasons, you may devise a set of unit tests to check the
step count data, i.e.~check the maximum and minimum step count, check
the difference between each day.

\begin{itemize}
\tightlist
\item
  If the daily count looks like c(4000, 5000, 5500, 5500, 20000), the
  maximum check will flag the data for investigate the maximum. The
  difference between days test will also flag the data
\item
  If the daily count looks like c(20000, 20000, 20000, 20000, 20000),
  the maximum check will flag the data for investigate the maximum.
\end{itemize}

Some part of the space is impossible: c(0, 4000, 5000, 5500, 5500) is
flagged by the minimal tests but won't cause an average of 8000 average
step.

The statistical procedure of averaging 5 numbers ``around 5000'' to get
a mean of 5000 is \emph{consistent} meaning if all the numbers are
around 5000, we are guaranteed to get a mean around 5000. We could
devise 5 unit tests to check each number. Since you're more familiar
with your daily life, you may realise the step count may increase since
you were in a hiking trip in the last two days. This may prompt you to
check the step count.

In a data analysis, it is not practical to check every entry of the
data, a similar strategy of devising tests to check for

\begin{itemize}
\tightlist
\item
  The combination of unit tests are not unique
\item
  The unit tests provide guidance for diagnosing the results, but are
  not red flags: c(2000, 2000, 5000, 8000, 8000) will likely to fail the
  max diff test but receive a within expectation mean.
\end{itemize}

\section{Applications}\label{sec-examples}

@@

### 2024-09-12 index.qmd
the toy example + some notesg

[background]

@@

* describe/ define what analysis plan is, expectation (outcome and plan)

* why it is useful to be explicit about expectation?

    * expectations can be used to formulate unit tests, which helps to divide the "result universe"

# Examples {#sec-examples}

@@

(these two paragraphs may go to a new section *discussion*) A result universe constructed in this example can be presented to analysts to answer the what-if questions raised in the data analysis. What if the sample size is increased? What if temperature is added to the model? What would the results expect to be changed when the correlation structure is different? For analysts, expectations can be used as unit tests to divide the result universe, which allows them to understand the impact of each factor on the outcome expectation.

The result universe also provides a holistic view of how the results obtained by the analysts are situated in all possible results. This can be seen as a direct towards trustworthy data analysis, where the audience of the analysis to exercise their own cognitive model [@grolemund_cognitive_2014] to evaluate the results reported.

>>>>

[background - In a data analysis workflow]

For expert analysts, the expectation of the analysis process is often implicit. They may have a mental model of the analysis process and the expected results. However, for non-expert analysts, the expectation of the analysis process may not be clear. They may not know what to expect from the analysis process, or what to look for when an analysis process fails the expectation. An analysis plan can help to make the expectation of the analysis process explicit. It can help to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process.

@@

* describe/ define what analysis plan is, expectation (outcome and plan)
    * expectations can be used to formulate unit tests, they can be used to guide to diagnose in the analysis process: data issue, unreasonable expectation -- expectation is "wrong" 
    * some expectations can be about data: e.g. check temporal gaps in time series data. Some others are statistical checks: sample size, outliers, model specification, or checks based on domain knowledge: include temperature in the mortality ~ PM10 model.

Q: Whether we should formulate these concept with math notation?? A: only if it helps

An analysis plan is a set of analysis steps plus their expected outcomes. An analysis plan consists of two parts: outcome expectation and plan expectation. Outcome expectation is the expected result of the analysis, for example, a significant p-value or a coefficient estimate within a certain range. Plan expectation is the expected results from the intermediate steps of the analysis, for example, the expected distribution of the data, or the expected correlation structure between variables.

While there is typically one major result expectation in an analysis, multiple plan expectations can be formulated as unit tests to guide the analysis process. These plan expectations serve as road marks to test, at various stages of the analysis, whether the results are within expectation. When the outcome is out of expectation, these plan expectations can serve as unit tests to diagnose the analysis process.

* why it is useful to be explicit about expectation? 
    * with simulation, the unit tests can be used to construct the "result universe" 
    * With formulated expectation, one can generate multiple version of the dataset for teaching statistics

When the expectation is formulated, we can also construct values fulfill or not fulfill the expectation. With simulation, this allows to construct the set of all possible results that can be obtained from the analysis process, which we called the result universe. With such a result universe, we can see how the results obtained by the analysts are situated in all possible results. When accompanied by the result universe, it is clearer how different factors in the analysis affects the main results. This can be seen as a direct towards trustworthy data analysis, where the audience of the analysis to exercise their own cognitive model [@grolemund_cognitive_2014] to evaluate the results reported.

Let's think about a 5-day step count. You make a resolution to walk on average 5000 steps a day (your expectation) and using an app to record your step count. After 5 days, the app tells you've walked on average 8000 steps.

It is easy to come up with reasons why an 8000 average step is resulted based on common sense:

1. you may run a 10k on day 1, resulting a high step count on the day (outlier on the right).
2. you left your phone at home on day 3, resulting a zero or minimal step count on the day (outlier on the left).
3. you may realise the step count may increase since you were in a hiking trip in the last five days (average shift).

Based on these reasons, you may devise a set of unit tests to check the step count data, i.e. check the maximum and minimum step count, check the difference between each day.

- If the daily count looks like c(4000, 5000, 5500, 5500, 20000), the maximum check will flag the data for investigate the maximum. The difference between days test will also flag the data
- If the daily count looks like c(20000, 20000, 20000, 20000, 20000), the maximum check will flag the data for investigate the maximum.

Some part of the space is impossible: c(0, 4000, 5000, 5500, 5500) is flagged by the minimal tests but won't cause an average of 8000 average step.

The statistical procedure of averaging 5 numbers "around 5000" to get a mean of 5000 is *consistent* meaning if all the numbers are around 5000, we are guaranteed to get a mean around 5000. We could devise 5 unit tests to check each number. Since you're more familiar with your daily life, you may realise the step count may increase since you were in a hiking trip in the last two days. This may prompt you to check the step count.

In a data analysis, it is not practical to check every entry of the data, a similar strategy of devising tests to check for

- The combination of unit tests are not unique
- The unit tests provide guidance for diagnosing the results, but are not red flags: c(2000, 2000, 5000, 8000, 8000) will likely to fail the max diff test but receive a within expectation mean.

# Applications {#sec-examples}

@@
